---
title: "Final Project"
author: "Oilivia Wagner| Weijia Xiong | Wurongyan Zhang | Yiling Yang "
output: 
  pdf_document:
    latex_engine: xelatex
    toc: true
    toc_depth: 4
mainfont: Times New Roman
fontsize: 11pt
geometry: "margin=1in"
header-includes:
    - \usepackage{setspace}
    - \doublespacing
---

```{r setup, include=FALSE}
library(tidyverse)
library(readr)
library(readxl)
library(faraway)
library(broom)
library(dplyr)
library(kableExtra)
library(knitr)
library(modelr)
library(MASS)
library(arsenal)
library(psych)
library(gridExtra)
library(corrplot)

knitr::opts_chunk$set(
	echo = FALSE,
	message = FALSE,
	warning = FALSE,
  fig.width = 10, 
	fig.height = 6,
	fig.align = "center",
	out.width = "80%"
)


kable  =  function(data) {
    knitr::kable(
    data,
    align = rep('c', 15),
    longtable = F, 
    booktabs = T, 
    escape = T,
    digits = 4
  ) %>% 
  kable_styling(
    latex_options = c("striped","hold_position","scale_down","repeat_header"),
    font_size = 10
  )
}
```


```{r}
## load and tidy the data
lawsuit = read_csv("./data/lawsuit.csv") %>% 
  janitor::clean_names() %>% 
  mutate(
    dept = recode(dept, 
                  "1" = "Biochemistry",
                  "2" = "Physiology",
                  "3" = "Genetics",
                  "4" = "Pediatrics",
                  "5" = "Medicine",
                  "6" = "Surgery") %>% as.factor(),
    gender = recode(gender,
                    "1" = "Male", 
                    "0" = "Female") %>% as.factor(),
    clin = recode(clin, 
                  "1" = "Clinical",
                  "0" = "Research") %>% as.factor(),
    cert = recode(cert, 
                  "1" = "Board certified", 
                  "0" = "Not certified") %>% as.factor(),
    rank = recode(rank, 
                  "1" = "Assistant",
                  "2" = "Associate", 
                  "3" = "Full professor") %>% as.factor()
  )

lawsuit = 
  lawsuit %>% 
  mutate(
  sal9495 = log(sal94 + sal95)/2,
  sal_original = (sal94 + sal95)/2
  )
```

\pagebreak
## Abstract

## Introduction

In our project, we examined the whether gender discrimination existed in setting salaries for people in the academia or higher education institutions.

## Exploratory data analysis

Data used in our studies were collected from 261 individuals who work in the academia or higher education institutions. The raw dataset contains following features. 

- **Dept**: 1= Biochemistry/Molecular Biology 2= Physiology 3= Genetics 4= Pediatrics 5= Medicine 6= Surgery 
- **Gender**: 1= Male, 0= Female
- **Clin**: 1= Primarily clinical emphasis, 0= Primarily research emphasis
- **Cert**: 1= Board certified, 0= not certified
- **Prate**: Publication rate (# publications on cv) / (# years between CV date and MD date) 
- **Exper**: # years since obtaining MD
- **Rank**: 1= Assistant, 2= Associate, 3= Full professor (a proxy for productivity)
- **Sal94**: Salary in academic year 1994
- **Sal95**: Salary after increment to Sal94

Table 1 contains the summary of variables in the dataset. Fortunately, there are no missing values in our dataset. We then need to examine each interested variable against the main effect and main interest.

**Salary**
```{r}
model_orgin = lm(sal_original ~ gender + dept + clin + cert + prate + exper + rank, data = lawsuit)
boxcox(model_orgin)
```
Since $\lambda = 0$, we use log transformation.

```{r}
p1 = 
  lawsuit %>% 
  ggplot(aes(x = sal_original, y = ..density..)) +
  geom_histogram(binwidth = 10000) +
  geom_density(alpha = 0.3,fill = "red") +
  labs(title = 'Salary Distribution')


p2 =
  lawsuit %>% 
  ggplot(aes(x = sal9495, y = ..density..)) +
  geom_histogram(binwidth = 0.05) +
  geom_density(alpha = 0.3,fill = "red") +
  labs(title = 'Salary Distribution(After log transformation)')

grid.arrange(p1,p2,ncol = 2)

```


**Department**

Fig 1.1. shows that the gender ratios in the department of Genetics and Physiology are very balanced, compared with department of Medicine  and Surgery, which are very imbalanced. The differences in department of Biochemistry and Pediatrics are moderate. Fig 1.2 includes our main interest the salary into account. It seems that across all department, male earn more than female do. However, before further analysis, we cannot tell whether those difference are significant.

```{r}
p1 <- ggplot(aes(x=dept,fill=gender),data=lawsuit)+
  geom_bar(position=position_dodge())+
  geom_text(stat = 'count',vjust=1.6, position = position_dodge(0.9),aes(x = dept, label = stat(count)))+
  ggtitle("Fig.1.1 Gender Distribution of Department")+
  theme_minimal()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  xlab("Department") +
  ylim(0,60)+
  scale_fill_brewer(palette="Greens")


p2 <- ggplot(aes(x=dept,y=sal9495,fill=gender),data=lawsuit) +
  geom_boxplot()+
  theme_minimal()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("Fig.1.2 Average Salary across Department & Gender") +
  ylab("Average Salary")+
  xlab("Department")

grid.arrange(p1,p2,ncol=2)
```

**Clinical or Research Division**

Fig 2.1 shows that in either division, there are more male than female, especially the clinical division. Regarding salar, again, male in either division earned more than female do.

```{r}
p1 <- ggplot(aes(x=clin,fill=gender),data=lawsuit)+
  geom_bar(position=position_dodge())+
  geom_text(stat = 'count',vjust=1.6, position = position_dodge(0.9),aes(x = clin, label = stat(count)))+
  ggtitle("Fig.2.1 Gender Distribution across Division")+
  theme_minimal()+
  xlab("") +
  ylim(0,120)+
  scale_fill_brewer(palette="Greens")

p2 <- ggplot(aes(x=clin,y=sal9495,fill=gender),data=lawsuit) +
  geom_boxplot()+
  theme_minimal()+
  ggtitle("Fig.2.2 Average Salary across clin & gender") +
  ylab("Average Salary")+
  xlab("")

grid.arrange(p1,p2,ncol=2)
```

**Certfication status**

Fig.3. shows that the amount of certified male outnumbers surely the amount of certified female. However, for those without certification, theose are just even. For the salar, male again earn more than female do regardless of his or her certification status.


```{r}
p1 <- ggplot(aes(x=cert,fill=gender),data=lawsuit)+
  geom_bar(position=position_dodge())+
  geom_text(stat = 'count',vjust=1.6, position = position_dodge(0.9),aes(x = cert, label = stat(count)))+
  ggtitle("Fig.3.1 Gender Distribution against Certification Status")+
  theme_minimal()+
  xlab("") +
  ylim(0,120)+
  scale_fill_brewer(palette="Greens")

p2 <- ggplot(aes(x=cert,y=sal9495,fill=gender),data=lawsuit) +
  geom_boxplot()+
  theme_minimal()+
  ggtitle("Fig.3.2 Average Salary across Cert & Gender") +
  ylab("Average Salary")+
  xlab("")

grid.arrange(p1,p2,ncol=2)
```


**Prate**

Fig 4.1 shows the density plot of publication rate for both male and female. No obvious difference was observed. Fig 4.2 implies that there might be linear trend between average salary and publication rate.However, it's hard to tell whetehr there is any difference on the effect on the salar regarding the gender.s 

```{r}
p1 <- ggplot(data = lawsuit,aes(x=prate,color=gender,fill=gender))+
  geom_density(linetype="dashed",alpha=0.4)+
  ggtitle("Fig.4.1 Density of Publication Rate against Gender")+
  ylab("")+
  xlab("Publication Rate")


p2 <- ggplot(data = lawsuit,aes(x=prate,y=sal9495,color=gender))+
  geom_point()+
  ggtitle("Fig.4.2 Average Salary against Prate across Gender")+
  xlab("Prate")+
  ylab("Average Salary")

grid.arrange(p1,p2,ncol=2)
```

**Exper**

Fig.5.1 shows that the densities of experience year for both male and female are both very skewed, and seems that female have a more skewed trend. Although there are not obvious linear trend between salar and experience year, the salar for male spread more widly than those for female and individual with high salar are dominately male.

```{r}
#geom_vline(data = prate_mean,aes(xintercept=mean,color=gender,linetype="dashed",size=0.5))+

p1 <- ggplot(data = lawsuit,aes(x=exper,color=gender,fill=gender))+
  geom_density(linetype="dashed",alpha=0.4)+
  ggtitle("Fig 5.1 Density of experience Year against Gender")+
  ylab("")+
  xlab("Experience Year")


p2 <- ggplot(data = lawsuit,aes(x=exper,y=sal9495,color=gender))+
  geom_point()+
  ggtitle("Fig 5.2 Average Salary against Experience Year across Gender")+
  xlab("Experience Year")+
  ylab("Average Salary")

grid.arrange(p1,p2,ncol=2)
```

**Rank**

Fig 6.1 shows that there are many female assistant professor and less associate and full professor  than male. Without surprising, in either rank, male earn more than female.

```{r}
p1 <- ggplot(aes(x=rank,fill=gender),data=lawsuit)+
  geom_bar(position=position_dodge())+
  geom_text(stat = 'count',vjust=1.6, position = position_dodge(0.9),aes(x = rank, label = stat(count)))+
  ggtitle("Fig.6.1 Gender Distribution against Rank")+
  theme_minimal()+
  xlab("") +
  ylim(0,120)+
  scale_fill_brewer(palette="Greens")

p2 <- ggplot(aes(x=rank,y=sal9495,fill=gender),data=lawsuit) +
  geom_boxplot()+
  theme_minimal()+
  ggtitle("Fig.6.2 Average Salary across Rank & Gender") +
  ylab("Average Salary")+
  xlab("")

grid.arrange(p1,p2,ncol=2)
```

Lastly, after examing the correlation matrix, we can see that there are some highly related variables  ($r > 80\%$). They are 1. Department (Dept) and Publish Rate (Prate) and 2. Clin and Prate. Some have morderate correlation such as 1. Experience Year (Exper) annd Cert (Certification status) and 2.Department (Dept) and Clin.

```{r}
res <- cor(read_csv("./data/lawsuit.csv")[,-c(1,9,10)])
corrplot(res, method = "number",type = "upper", order = "hclust", 
         tl.col = "black", tl.srt = 45)
```

Those imply potential collinearities. As we go through those variables, there are some outliers. In later section, we will examine further about them.

## Method

```{r}
reg1 = lm(sal9495 ~ gender, data = lawsuit)
summary(reg1)
```


```{r}
fit1 = lm(sal9495 ~ gender + dept, data = lawsuit)
summary(fit1)
anova(reg1,fit1)
```

```{r}
fit2 = lm(sal9495 ~ gender + clin, data = lawsuit)
summary(fit2)
anova(reg1,fit2)
```

```{r}
fit3 = lm(sal9495 ~ gender + cert, data = lawsuit)
summary(fit3)
anova(reg1,fit3)
```

```{r}
fit4 = lm(sal9495 ~ gender + exper, data = lawsuit)
summary(fit4)
anova(reg1,fit4)
```

```{r}
fit5 = lm(sal9495 ~ gender + rank, data = lawsuit)
summary(fit5)
anova(reg1,fit5)
```
There is no need to consider prate since the correlation of prate and dept is -0.87. Otherwise there might be collinearity.

From linear regression results, rank is not signifiant. From partial F test, p is more than 0.05, which shows that model with rank is not superior.

Since the changes of estimate of genderMale are more than 10% and all p-values from partial F test are less than 0.05, dept, clin, cert and exper are all confounders. So we need to add these covariates into the model.



First we need to add department....
```{r}
reg2 = lm(sal9495 ~ gender + dept, data = lawsuit)
summary(reg2)
```

```{r}
## add clin
fit_gdclin = lm(sal9495 ~ gender + dept + clin, data = lawsuit)
summary(fit_gdclin)
anova(reg2,fit_gdclin)
```


```{r}
## add cert
fit_gdcert = lm(sal9495 ~ gender + dept + cert, data = lawsuit)
summary(fit_gdcert)
anova(reg2,fit_gdcert)
```


```{r}
## add exper
fit_gde = lm(sal9495 ~ gender + dept + exper, data = lawsuit)
summary(fit_gde)
anova(reg2,fit_gde)
```

```{r}
## add rank
fit_gdr = lm(sal9495 ~ gender + dept + rank, data = lawsuit)
summary(fit_gdr)
anova(reg2,fit_gdr)
```



Result: Compared with other covariates, F value of adding exper is the largest. So we add exper into our model.

```{r}
reg3 = lm(sal9495 ~ gender + dept + exper, data = lawsuit)
summary(reg3)
```

Then we try to look for next covariate.
```{r}
## add clin
fit_gdeclin = lm(sal9495 ~ gender + dept + exper + clin, data = lawsuit)
summary(fit_gdeclin)
anova(reg3,fit_gdeclin)
```

```{r}
## add cert
fit_gdecert = lm(sal9495 ~ gender + dept + exper + cert, data = lawsuit)
summary(fit_gdecert)
anova(reg3,fit_gdecert)
```

```{r}
## add rank
fit_gderank = lm(sal9495 ~ gender + dept + exper + rank, data = lawsuit)
summary(fit_gderank)
anova(reg3,fit_gderank)
```

Result: Compared with other covariates, F value of adding clin is the largest. So we add exper into our model.

```{r}
reg4 = lm(sal9495 ~ gender + dept + exper + clin, data = lawsuit)
summary(reg4)
```

Then we try to look for next covariate :
 
```{r}
## add cert
fit_gdeclin_cert = lm(sal9495 ~ gender + dept + exper + clin + cert, data = lawsuit)
summary(fit_gdeclin_cert)
anova(reg4,fit_gdeclin_cert)
```


```{r}
## add rank
fit_gdeclin_rank = lm(sal9495 ~ gender + dept + exper + clin + rank, data = lawsuit)
summary(fit_gdeclin_rank)
anova(reg4,fit_gdeclin_rank)
```
Result: Compared with other covariates, F value of adding cert is the largest. So we add cert into our model.

```{r}
reg5 =  lm(sal9495 ~ gender + dept + exper + clin + cert, data = lawsuit)
summary(reg5)
```


```{r}
## add rank
fit_gdeclin_cert_rank = lm(sal9495 ~ gender + dept + exper + clin + cert + rank, data = lawsuit)
summary(fit_gdeclin_cert_rank)
anova(reg5,fit_gdeclin_cert_rank)
```


Result:

p value of partial F test is more than 0.05, which shows that the model adding cert is superior.
Therefore, we need to add cert in the model.


```{r}
final_main_model_rank = 
  lm(sal9495 ~ gender + dept + exper + cert + clin + rank, data = lawsuit)
summary(final_main_model_rank)
```

## add rank main effect  + interaction 
```{r}
fit_inter_1_rank = lm(sal9495 ~ gender * dept +  exper  + clin + cert + rank, data = lawsuit)
summary(fit_inter_1_rank)
anova(final_main_model_rank,fit_inter_1_rank)
```

not significant

```{r}
fit_inter_2_rank = lm(sal9495 ~ gender * exper +  dept  + clin + cert + rank, data = lawsuit)
summary(fit_inter_2_rank)
anova(final_main_model_rank,fit_inter_2_rank)
```

significant

```{r}
fit_inter_3_rank = lm(sal9495 ~ gender * clin +  dept  + exper + cert + rank, data = lawsuit)
summary(fit_inter_3_rank)
anova(final_main_model_rank,fit_inter_3_rank)
```

not significant


```{r}
fit_inter_4_rank = lm(sal9495 ~ gender * cert +  dept  + exper + clin + rank, data = lawsuit)
summary(fit_inter_4_rank)
anova(final_main_model_rank,fit_inter_4_rank)
```
not significant


```{r}
fit_inter_5_rank = lm(sal9495 ~ gender * rank +  dept  + exper + clin + cert, data = lawsuit)
summary(fit_inter_5_rank)
anova(final_main_model_rank,fit_inter_5_rank)
```
significant.



----------------------------------------
**Final Model with Interaction:**
```{r}
final_model = lm(sal9495 ~ gender * rank + exper+ dept + clin + cert + rank, data = lawsuit)
summary(final_model)

```

```{r}
anova(final_model)
```

**Remove outlier**
```{r}
influence.measures(final_model) %>% summary() %>% tidy()
### 184 outlier
newlawsuit = lawsuit[-184,]
model_ge_without184 = lm(sal9495 ~ gender * rank + exper+ dept + clin + cert + rank, data = lawsuit)

summary(model_ge_without184)
anova(model_ge_without184)
par(mfrow = c(2, 2))
plot(model_ge_without184)
```


Stratification: 

## on rank
```{r}
# Associate
associate_fit = lawsuit %>% 
  filter(rank == "Associate") %>% 
  lm(sal9495 ~ gender+dept+clin+cert+exper, data = .)
summary(associate_fit)
anova(associate_fit)

# Assistant
assistant_fit = lawsuit %>% 
  filter(rank == "Assistant") %>% 
  lm(sal9495 ~ gender+dept+clin+cert+exper, data = .)
summary(assistant_fit)
anova(assistant_fit)

# Professor
prof_fit = lawsuit %>% 
  filter(rank == "Full professor") %>%
  lm(sal9495 ~ gender+dept+clin+cert+exper, data = .)
summary(prof_fit)
anova(prof_fit)
```

Gender is only significant in assistant professors' salary.

## Conclusions

## Discussion

\pagebreak
## Figures and tables


```{r, results = "asis"}
## classified by gender
my_controls = tableby.control(
               total = T,
               test = T,  # No test p-values yet
               numeric.stats = c("meansd", "medianq1q3", "range", "Nmiss2"),
               cat.stats = c("countpct", "Nmiss2"),
               stats.labels = list(
                 meansd = "Mean (SD)",
                 medianq1q3 = "Median (IQR)",
                 range = "Min - Max",
                 Nmiss2 = "Missing",
                 countpct = "N (%)"))

tab2 = tableby( gender ~ dept+clin+cert+prate + exper + rank+sal94 + sal95, 
                data = lawsuit, control = my_controls)

my_labels = list(dept= "Department",
                 clin = "Clinical",
                 cert = "Certified",
                 prate = "Publication Rate",
                 exper = "Years since obtaining MD",
                 rank = "Rank",
                 sal94 = "Salary in academic year 1994",
                 sal95 = "Salary after increment to Sal94",
                 gender = "Gender(%)")

summary(tab2, title = "Table 1:  Summary of All Variables", 
        labelTranslations = my_labels, 
        text = T) %>% 
  knitr::kable(
    align = rep('c', 15),
    longtable = T, 
    booktabs = T, 
    escape = T,
    digits = 2
  ) %>% 
  kable_styling(
    latex_options = c("striped", "hold_position","repeat_header","scale_down"),
    full_width = T ,
    font_size = 10
  )
```

\pagebreak
## References

\pagebreak
## Appendix
